Supplement to ``{\mytitle}''

### 2023-08-23 si/si.Rnw
Updated SI w/ haitipkg, minor edits

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}

@@

Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for our purposes.
For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
However, for applicability to larger populations, we use instead the following Euler scheme.
Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
\begin{equation}
\label{eq:numerics1}
\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
\end{equation}
In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
Then, for $i\neq \demography$ and $j\neq i$, and writing
\begin{equation}
\label{eq:numerics2}
\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
\end{equation}
we calculate transition probabilities
\begin{eqnarray}
\label{eq:numerics3}
p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
\\
\label{eq:numerics4}
p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
\end{eqnarray}
These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.

@@

Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
In general, we used the same measurement models that were reported by \cite{lee20}.

@@

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
\begin{equation}
  \label{model1:meas}
  Y_{n} \mid \Delta N_{EI}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
\end{equation}
where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z \big(1 + \frac{\reportRate z}{\obsOverdispersion}\big)$.

@@

As mentioned in the main text, Model~2 was fit using reported case counts that were transformed using the natural logarithm.
We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases as the loss function.
\begin{equation}
  \label{model2:meas}
  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\log \big(\reportRate z(n) + 1\big), \obsOverdispersion^2 \right),
\end{equation}
where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $n-1 \in 0:N-1$.
Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
\begin{equation}
  \label{model2:measEpi}
  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
\end{equation}
which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:measEnd}
  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate \big(z_1(n) + z_2(n)\big), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.

@@

In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
\begin{equation}
  \label{model3:meas}
  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(n) \sim \text{NB}\left(\reportRate z_{u}(n), \obsOverdispersion \right),
\end{equation}
where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $n-1 \in 0:N-1$.

@@

To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values could be known, or could be a non-random function of the unknown parameter vector, $\theta$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

@@

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.

@@

Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.

@@

As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.
The remaining hidden states are estimated by enforcing model dynamics using this result.

% \section{Block Panel Iterated Filter}
%
% The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
% Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
% Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.
%
% In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{u}$, for units $u \in \seq{1}{U}$.
% Here, we say that a parameter $\psi_{u}$ for $u \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $k \neq u \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
% The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{1:U})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
% Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.
% This is because it is likely that, when updating the parameters for a given unit, several parameter vector $\theta$ particles will not be resampled due to having low weights.
% When the number of units is large, this will lead to a decreasing number of ancestral particles to sample unit-specific parameters $\psi_k$ as $k$ increases.
%
% As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
% In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
% The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
% unit are being considered.
% We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.
% No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
%
% \begin{algorithm}[ht]
%    \caption[BPIF]{\textbf{BPIF}. \\
%     {\bf Inputs}:\\\hspace{\textwidth}
%     Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $\seq{1}{U}$.\\\hspace{\textwidth}
%     Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
%     Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
%     Data $y_{u, n}^*$,for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
%     Number of iterations, $M$.\\\hspace{\textwidth}
%     Number of particles, $J$.\\\hspace{\textwidth}
%     Starting parameter swarm, $\Theta_j^0 = \big(\Phi_j^0, \Psi_{1:U, j}^0\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.\\\hspace{\textwidth}
%     Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u \in \seq{1}{U}$, $n \in 0:N_u$.\\\hspace{\textwidth}
%     Perturbation sequence, $\sigma_m$ for $m \in 1:M$.\\\hspace{\textwidth}
%     {\bf Output:} \\\hspace{\textwidth}
%     Final parameter swarm, $\Theta_{j}^{m} = \big(\Phi_j^m, \Psi_{\seq{1}{U}, j}^m\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.
%     \label{alg:bpif}}
% %\BlankLine
% \For{$m \in 1:M$}{
%   Set $\Phi_{0, j}^m = \Phi_j^{m-1}$ for $j \in \seq{1}{J}$\;
%     \For{$u \in \seq{1}{U}$}{
%       Set $\Phi_{u, 0, j}^{F, m} \sim h_{u, 0}(\phi | \Phi_{0, j}^m; \sigma_m)$ for $j \in \seq{1}{J}$\;
%       Set $\Psi_{u, 0, j}^{F, m} \sim h_{u, 0}(\psi | \Psi_{u, j}^{m - 1}; \sigma_m)$ for $j \in \seq{1}{J}$\;
%       Set $\Theta_{u, 0, j}^{F, m} = \big(\Phi_{u, 0, j}^{F, m}, \Psi_{u, 0, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
%       Initialize $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in \seq{1}{J}$\;
%         \For{$n \in \seq{1}{N_u}$} {
%           Set $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in \seq{1}{J}$\;
%           $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
%           $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
%           Draw $k_{\seq{1}{j}}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in \seq{1}{J}$\;
%           Set $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m} = \big(\Phi_{u, n, k_j}^{P, m}, \Psi_{u, n, k_j}^{P, m}\big)$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in \seq{1}{J}$\;
%         }
%       Set $\Theta_{u, j}^m = \big(\Phi_{u, N_u, j}^{F, m}, \Psi_{u, N_u, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
%     }
%   Set $\Theta_j^m = \big(\Phi_{U, j}^m, \Psi_{1:U, j}^m\big)$ for $j \in \seq{1}{J}$\;
% }
% \end{algorithm}

@@

# This function sets the rw.sd for named parameters, first search.
set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.01)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

# This function sets the rw.sd for named parameters, second search.
set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.003)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

# This function sets the rw.sd for named parameters, final search.
set_rw_3 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.0025)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.0025)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.0025)
  } else {
    return(0)
  }
}

# First Search RW
reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
names(reg_rw_1.sd) <- est_param_names_expanded
chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)

# Second Search RW
reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
names(reg_rw_2.sd) <- est_param_names_expanded
chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)

# Third Search RW
reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
names(reg_rw_3.sd) <- est_param_names_expanded
chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)

@@

# Fit the model and save the results
haiti3_orig_fit <- bake(
  file = save_file, {
    fit_coupled_haiti3(
      nsearches = N_SEARCHES,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search3 = SEARCH3,
      search_rho = FALSE,
      search_gamma = FALSE,
      ncores = cores
    )
  }
)

@@

# Create dataset that will be used to fit ARMA models. To do this, we
# need to make sure that missing values are marked as NA, so we check
# for any jumps in time.
MODEL3_CASES <- haitiCholera %>%
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) %>%
  dplyr::mutate(date = as.Date(date)) %>%
  dplyr::select(-report) %>%
  dplyr::mutate(time_diff = date - lag(date))

>>>>

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\Iinit{I_{0,0}}
\newcommand\Einit{E_{0, 0}}
\newcommand\reportChange{\delta}

@@

Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
However, for applicability to larger populations, we use instead the following Euler scheme.
Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
\begin{equation}
\label{eq:numerics1}
\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
\end{equation}
In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
Then, for $i\neq \demography$ and $j\neq i$, and writing
\begin{equation}
\label{eq:numerics2}
\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
\end{equation}
we calculate transition probabilities
\begin{eqnarray}
\label{eq:numerics3}
p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
\\
\label{eq:numerics4}
p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
\end{eqnarray}
These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.

@@

Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.

@@

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
\begin{equation}
  \label{model1:meas}
  Y_{n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
\end{equation}
where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$, and $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.

@@

As mentioned in the main text, Model~2 was fit using reported case counts that were transformed using the natural logarithm.
We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases as the loss function.
\begin{equation}
  \label{model2:meas}
  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\log \big(\reportRate \reportChange + 1\big), \obsOverdispersion^2 \right),
\end{equation}
where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
The measurement model they used for the epidemic phase is
\begin{equation}
  \label{model2:measEpi}
  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\reportRate \reportChange, \obsOverdispersion^2 \right),
\end{equation}
% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:measEnd}
  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange_1, \Delta N_{EA}(n) = \reportChange_2 \sim \text{N}\left(\reportRate \big(\reportChange_1 + \reportChange_2\big), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.

@@

In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
\begin{equation}
  \label{model3:meas}
  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
\end{equation}
where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $t_{n-1}$.

@@

To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

@@

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

@@

Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

@@

The latent states of this model are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.
The remaining hidden states are estimated by enforcing model dynamics using this result.
% TODO: Check this section.

@@

# This function sets the rw_sd for named parameters, first search.
set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.01)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

# This function sets the rw_sd for named parameters, second search.
set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.003)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

# This function sets the rw_sd for named parameters, final search.
set_rw_3 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.0025)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.0025)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.0025)
  } else {
    return(0)
  }
}

# First Search RW
reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
names(reg_rw_1.sd) <- est_param_names_expanded
chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)

# Second Search RW
reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
names(reg_rw_2.sd) <- est_param_names_expanded
chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)

# Third Search RW
reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
names(reg_rw_3.sd) <- est_param_names_expanded
chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)

@@

# Fit the model and save the results
# TODO: Remove SEARCH3 when re-running everything.
haiti3_orig_fit <- bake(
  file = save_file, {
    fit_coupled_haiti3(
      nsearches = N_SEARCHES,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search3 = SEARCH3,
      search_rho = FALSE,
      search_gamma = FALSE,
      ncores = cores
    )
  }
)

@@

MODEL3_CASES <- haitiCholera |>
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::select(-report) |>
  dplyr::mutate(time_diff = date - lag(date)) |>
  dplyr::filter(date >= "2010-11-20")
